\bf Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-06-12 pkg/paper/Paper/JAE.Rnw
Renamed "draft" to "report", deleted .pdf.
New JAE.Rnw with mimicked JAE style and increased baselinestretch.

Touch-ups in .bib.



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@120 b679b2f7-6539-45d0-a605-5ddcb976cde9

\documentclass[nojss]{jss}
\usepackage{amsfonts,amstext,amsmath}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

@@

%% for internal use
\newcommand{\fixme}[1]{\emph{\marginpar{FIXME} (#1)}}
\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}

\title{Structural Breaks in Inflation Dynamics within the European Monetary Union}

\author{Thomas Windberger\\Universit\"at Innsbruck \And
        Achim Zeileis\\Universit\"at Innsbruck}
%% \author{\hfill Thomas Windberger \hfill Achim Zeileis\thanks{Correspondence
%% to: Achim Zeileis, Department of Statistics,  Universit\"at Innsbruck,
%% Universit\"atsstr.~15, A-6020 Innsbruck, Austria. Tel: +43/512/507-7103. Fax: +43/512/507-2851.
%% E-mail: \email{Achim.Zeileis@R-project.org}} \hfill \hfill \\
%% {\small \it Department of Statistics, Universit\"at Innsbruck, Austria}}
\Plainauthor{Thomas Windberger, Achim Zeileis}

\renewcommand{\abstractname}{Summary}
\Abstract{
  To assess the effects of the EMU on inflation rate dynamics of
  its member states, the inflation rate series for 21~European countries
  are investigated for structural changes. To capture changes in mean,
  variance, and skewness of inflation rates, a generalized logistic
  model is adopted and complemented with structural break tests and
  breakpoint estimation techniques. These reveal considerable differences
  in the patterns of inflation dynamics and the structural changes
  therein. Overall, there is a convergence towards a lower mean inflation rate
  with reduced skewness, but it is accompanied by an increase in variance.
}

\Keywords{inflation rate, structural break, EMU, generalized logistic distribution}

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstra{\ss}e 15\\
  A-6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
  Telephone: +43/12/507-7103 \\
  Fax: +43/512/507-2851 \\
}

@@

@@

\begin{table}[p!]
\begin{center}
\footnotesize
\begin{tabular}{llrrrlll}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline

@@

\caption{\label{tab:results1} Estimated breakpoints, fitted moments of generalized logistic
distribution per segment, and ERM information for all countries under investigation.}
\end{center}
\end{table}

@@

\begin{table}[t!]
\begin{center}
\begin{tabular}{llrrrr}
\hline
Moment & Quartile & \Sexpr{yrs[1]} & \Sexpr{yrs[2]} & \Sexpr{yrs[3]} & \Sexpr{yrs[4]}\\
\hline

@@

\end{tabular}
\caption{\label{tab:results2} Aggregation of estimated moments from Table~\ref{tab:results1} across countries in
Jan~\Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]}, and \Sexpr{yrs[4]}. Dots signal that the moments have not
changed from the previous time point, i.e., that no breakpoint was estimated in the corresponding time period.}
\end{center}
\end{table}

@@

\section*{Computational details}

@@

Replication scripts for all analyses are provided within the \pkg{glogis} package: An overview
is provided in \code{demo(package = "glogis")} and the scripts can be launched easily, e.g.,
via \code{demo("Austria", package = "glogis")} etc.

\section*{Acknowledgments}

We are grateful to Jesus Crespo Cuaresma for comments and discussions about
an earlier version of the manuscript, and also to the participants of the workshop
of the Research Platform ``Empirical and Experimental Economics'' for stimulating discussions,
leading to substantial improvements of the paper.

\bibliography{papers}

\end{document}
\SweaveOpts{eval=FALSE}

\clearpage

\begin{appendix}

\section{Data and sales periods}\label{sec:sales}

In addition to Germany, Portugal, Belgium, Italy, Luxembourg, and Spain, there
are a number of other countries where the sales periods are regulated by law.
In France the sale periods time span is regulated by each department. In Greece there
are also two set sales-periods in winter and summer. In Italy, the sales periods
are fixed each year by the Chambers of Commerce.  Portugal has two set sales
periods as well, but for two month (January and February as well as Mid-July to
Mid-September). The source for this information is \citet{konseur}. According to
\cite{pltm}, the official dates of the inclusion of these sales periods are:
January 1998 for Germany and Portugal,  January 1999 for Luxembourg and Ireland and 
January 2000 for Belgium,  while Italy and Spain introduced them in January 2001.

\clearpage
\section{Graphs}

Figures~\ref{fig:nochange}--\ref{fig:eastern} depict the data underlying the results
in Section~\ref{sec:results} along with the GL-based means from Table~\ref{tab:results1}.

<<graphs-code>>=
get_data <- function(i) {
  rval <- do.call("merge",
    lapply(i, function(j) results[[j]]$data))
  colnames(rval) <- names(results)[i]
  rval
}

get_fitted <- function(i) {
  rval <- do.call("merge",
    lapply(i, function(j) fitted(results[[j]]$breakpoints, breaks = results[[j]]$nbreaks, type = "mean")))
  colnames(rval) <- names(results)[i]
  rval
}

library("lattice")
get_xyplot <- function(i) {
  nam <- names(results)[i]
  nam[nam == "CzechRepublic"] <- "Czech Republic"
  xyplot(merge(get_data(i), get_fitted(i)),
    screen = rep(nam, 2),
    col = rep(c(1, 4), each = length(i))) 
}
@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[p!]
\begin{center}
<<graphs-nochange, fig=TRUE, height=6.3, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(2, 4, 8, 14)))
@
\caption{\label{fig:nochange} No change.}

<<graphs-phase1, fig=TRUE, height=4.9, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(1, 6, 7)))
@
\caption{\label{fig:phase1} Phase~1 convergence.}
\end{center}
\end{figure}

\begin{figure}[p!]
\begin{center}
<<graphs-phase2, fig=TRUE, height=6.3, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(9, 12, 13, 16)))
@
\caption{\label{fig:phase2} Phase~2 convergence.}

<<graphs-crisis, fig=TRUE, height=3.5, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(11, 19)))
@
\caption{\label{fig:crisis} Financial crisis.}
\end{center}
\end{figure}

\begin{figure}[p!]
\begin{center}
<<graphs-noneuro, fig=TRUE, height=3.5, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(20, 21)))
@
\caption{\label{fig:noneuro} Non-euro.}

<<graphs-eastern, fig=TRUE, height=9.1, width=7.5>>=
trellis.par.set(theme = standard.theme(color = FALSE))
print(get_xyplot(c(3, 5, 10, 15, 17, 18)))
@
\caption{\label{fig:eastern} Eastern countries.}
\end{center}
\end{figure}

\end{appendix}

>>>>

\documentclass[10pt,a4paper,twoside]{article}

%% additional packages
\usepackage{amsfonts,amstext,amsmath,a4wide,graphicx,color,thumbpdf}
\usepackage{hyperref}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

@@

%% BibTeX settings
\usepackage[authoryear,round]{natbib}
\bibliographystyle{jae}
\bibpunct{(}{)}{,}{a}{,}{,}
\newcommand{\doi}[1]{\href{http://dx.doi.org/#1}{\normalfont\texttt{doi:#1}}}

%% markup commands for code/software
\let\code=\texttt
\let\pkg=\textbf
\let\proglang=\textsf
\newcommand{\file}[1]{`\code{#1}'}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}

%% paragraph formatting
\renewcommand{\baselinestretch}{1.52}

%% mimic JAE style
\renewcommand{\section}{\secdef \mysec \mysecnn}
\newcommand{\mysec}[2][default]{\vspace{1.7\baselineskip}%
  \pdfbookmark[1]{#1}{Section.#1}%
  \refstepcounter{section}%
  \centerline{\large \thesection. \uppercase{#1}} \vspace{.5\baselineskip}}
\newcommand{\mysecnn}[1]{\vspace{1.7\baselineskip}%
  \centerline{\large #1} \vspace{.5\baselineskip}}
\renewcommand{\abstractname}{\normalfont SUMMARY}
\renewcommand{\refname}{REFERENCES}
\renewcommand{\thetable}{\Roman{table}}

%% page header (and currently no footer)
\usepackage{fancyhdr}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0pt}
\pagestyle{fancy}
\thispagestyle{plain}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[CE]{{\normalsize \uppercase{T.~Windberger and A.~Zeileis}}}
\fancyhead[CO]{{\normalsize \uppercase{Structural Breaks in Inflation Dynamics within the EMU}}} 
% \fancyfoot[LO,LE]{\small Copyright {\copyright} 20XX John Wiley \& Sons, Ltd.}
\fancypagestyle{plain}{
\fancyhf{}
% \fancyfoot[LO,LE]{\small This is a preprint of an article published in %
%   \textit{Journal of Applied Econometrics}, \textbf{XX}(X), XXX--XXX. \\%
%   Copyright {\copyright} 20XX John Wiley \& Sons, Ltd. \doi{10.1002/jae.XXXX} }
}

%% title information
\title{\bf Structural Breaks in Inflation Dynamics within the European Monetary Union}
\author{\hfill Thomas Windberger \hfill Achim Zeileis\thanks{Correspondence
to: Achim Zeileis, Department of Statistics,  Universit\"at Innsbruck,
Universit\"atsstr.~15, A-6020 Innsbruck, Austria. Tel: +43/512/507-7103. Fax: +43/512/507-2851.
E-mail: \email{Achim.Zeileis@R-project.org}} \hfill \hfill \\
{\small \it Department of Statistics, Universit\"at Innsbruck, Austria}}
\date{}

% hyperref setup
\definecolor{Red}{rgb}{0.5,0,0}
\definecolor{Blue}{rgb}{0,0,0.5}
\hypersetup{%
  pdftitle = {Structural Breaks in Inflation Dynamics within the European Monetary Union},
  pdfsubject = {},
  pdfkeywords = {inflation rate, structural break, EMU, generalized logistic distribution},
  pdfauthor = {Thomas Windberger, Achim Zeileis},
  %% change colorlinks to false for pretty printing
  colorlinks = {true},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

\maketitle

\begin{abstract}
  To assess the effects of the EMU on inflation rate dynamics of
  its member states, the inflation rate series for 21~European countries
  are investigated for structural changes. To capture changes in mean,
  variance, and skewness of inflation rates, a generalized logistic
  model is adopted and complemented with structural break tests and
  breakpoint estimation techniques. These reveal considerable differences
  in the patterns of inflation dynamics and the structural changes
  therein. Overall, there is a convergence towards a lower mean inflation rate
  with reduced skewness, but it is accompanied by an increase in variance.
\end{abstract}

\noindent {\bf Keywords:} inflation rate, structural break, EMU, generalized logistic distribution.

@@

\renewcommand{\baselinestretch}{1.15}
\begin{table}[p!]
\begin{center}
\footnotesize
\begin{tabular}{llrrrlll}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline

@@

\caption{\label{tab:results1} Estimated breakpoints, fitted moments of generalized logistic
distribution per segment, and ERM information for all countries under investigation.}
\end{center}
\end{table}
\renewcommand{\baselinestretch}{1.52}

@@

\renewcommand{\baselinestretch}{1.15}
\begin{table}[t!]
\begin{center}
\begin{tabular}{llrrrr}
\hline
Moment & Quartile & \Sexpr{yrs[1]} & \Sexpr{yrs[2]} & \Sexpr{yrs[3]} & \Sexpr{yrs[4]}\\
\hline

@@

\end{tabular}
\caption{\label{tab:results2} Aggregation of estimated moments from Table~\ref{tab:results1} across countries in
Jan~\Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]}, and \Sexpr{yrs[4]}. Dots signal that the moments have not
changed from the previous time point, i.e., that no breakpoint was estimated in the corresponding time period.}
\end{center}
\end{table}
\renewcommand{\baselinestretch}{1.52}

@@

\section*{COMPUTATIONAL DETAILS}

@@

Replication scripts for all analyses are provided within the \pkg{glogis} package: An overview
is provided in \code{demo(package = "glogis")} and the scripts can be launched easily, e.g.,
via\linebreak \code{demo("Austria", package = "glogis")} etc.

%% \section*{Acknowledgments}
%% 
%% We are grateful to Jesus Crespo Cuaresma for comments and discussions about
%% an earlier version of the manuscript, and also to the participants of the workshop
%% of the Research Platform ``Empirical and Experimental Economics'' for stimulating discussions,
%% leading to substantial improvements of the paper.

\bibliography{glogis}
